Packaging research artefacts with RO-Crate

### 2022-01-01 latex/ro-crate.tex
*ized -> ised (British spelling)

The conceptual notion of a \textit{Research Object}
\cite{doi:10.1016/j.future.2011.08.004} is thus realized with the RO-Crate
model and serialized using Linked Data constructs within the RO-Crate
metadata file.

@@

The possibilities of consuming\footnote{Some consideration is needed in processing of RO-Crates as
knowledge graphs, e.g. establishing absolute URIs for files inside a
ZIP archive, detailed in the RO-Crate specification: \url
{https://www.researchobject.org/ro-crate/1.1/appendix/relative-uris.html}.} RO-Crate metadata with such
powerful tools gives another strong reason for using Linked Data as a
foundation. This use of mature Web\footnote{Note that an RO-Crate is not required to be published on the
Web, see Section~\ref{sec:selfdescribed}.} technologies also means its
developers and consumers are not restricted to the Research Object
aspects that have already been specified by the RO-Crate community, but
can extend and integrate RO-Crate in multiple standardized ways.

@@

The RO-Crate specification provides a core set of conventions to
describe research outputs using types and properties applicable across
scientific domains. However we have found that domain-specific use of
RO-Crate will, implicitly or explicitly, form a specialized \textbf{profile}
of RO-Crate; i.e., {a set of conventions, types and properties that are
minimally required and one can expect to be present in that subset of
RO-Crates}. For instance, RO-Crates used for exchange of workflows will
have to contain a data entity of type \texttt{ComputationalWorkflow}, or
cultural heritage records should have a \texttt{contentLocation}.

@@

RO-Crate
mandates\footnote{\url{https://www.researchobject.org/ro-crate/1.1/appendix/jsonld.html}}
the use of flattened, compacted JSON-LD in the RO-Crate Metadata file
\texttt{ro-crate-metadata.json}\footnote{The avid reader may spot that the RO-Crate Metadata file use the
extension \texttt{.json} instead of \texttt{.jsonld}, this is to emphasize the
developer expectations as a JSON format, while the file's JSON-LD
nature is secondary. See \url
{https://github.com/ResearchObject/ro-crate/issues/82}.}
where a single \texttt{@graph} array contains all
the data and contextual entities in a flat list. An example can be seen
in the JSON-LD snippet in Listing \ref{lis1} below, describing a simple RO-Crate
containing data entities described using contextual entities.

@@

In JSON-LD, the \texttt{@context} is a reference to another JSON-LD document
that provides mapping from JSON keys to Linked Data term IRIs, and can
enable various JSON-LD directives to cater for customized JSON
structures for translating to RDF.

@@

The Pacific And Regional Archive for Digital Sources in Endangered
Cultures (PARADISEC\footnote{\url{https://www.paradisec.org.au/}}) \cite{doi:10125/4567}
maintains a repository of more than 500,000 files documenting
endangered languages across more than 16,000 items, collected and
digitized over many years by researchers interviewing and recording
native speakers across the region.

@@

This initially workflow-centric approach was further formalized as the
Wf4Ever Research Object Model \cite{doi:10.1016/j.websem.2015.01.003},
which is a general-purpose research artefact description framework.
This model is based on existing ontologies (FOAF, Dublin Core Terms,
OAI-ORE and AO/OAC precursors to the W3C Web Annotation Model
\cite{Ciccarese:17:WAD}) and adds specializations for workflow models and
executions using W3C PROV-O \cite{PROVO}. The Research Object statements
are saved in a \textit{manifest} (the OAI-ORE \textit{resource map}), with additional
annotation resources containing user-provided details such as title and
description.

We now claim that one barrier for wider adoption of the Wf4Eer Research
Object model for general packaging digital research artefacts was
exactly this re-use of multiple existing vocabularies (FAIR principle
I2: \textit{Metadata use vocabularies that follow FAIR principles}), which in
itself is recogized as a challenge \cite{doi:10.3233/978-1-61499-660-6-9}.
Adapters of the Wf4Ever RO model would have to navigate documentation
of multiple overlapping ontologies, in addition to facing the usual
Semantic Web development choices for RDF serialization formats,
identifier minting and publishing resources on the Web.

@@

\subsection{Example of formalized RO-Crate}

@@

A formalized RO-Crate in $\mathcal{L}_{\mathit{rocrate}}$ can be mapped to
different serializations.
Assume a simplified\footnote{This simplification does not cover the extensive list of literal
datatypes built-in to RDF 1.1, only strings and decimal real numbers.
Likewise, language of literals are not included.} language $\mathcal{L}_{\textit{RDF}}$
based on the RDF abstract syntax \cite{rdfworkinggroup_2014}:
%
{\fontsize{9.8}{11.8}\selectfont{\begin{eqnarray*}
\mathcal{L}_{\textit{RDF}} & \equiv& \bigl\{ \textit{Triple}(s,p,o),
\textit{IRI}(i), \textit{BlankNode}(b), \textit{Literal}(s), \mathbb{IRI},
\mathbb{S}, \mathbb{R} \bigr\}
\\
\mathbb{D}_{\textit{RDF}} & \equiv& \mathbb{S}
\\
\forall i \ \textit{IRI}(i) & \Rightarrow& i \in\mathbb{IRI}
\\
\forall s \forall p \forall o \ \textit{Triple}(s,p,o) & \Rightarrow& \bigl(
\textit{IRI}(s) \lor \textit{BlankNode}(s) \bigr) \land
\\
& & \textit{IRI}(p) \land
\\
& & \bigl(\textit{IRI}(o) \lor \textit{BlankNode}(o) \lor \textit{Literal}(o)
\bigr)
\\
\textit{Literal}(v) & \models& \textit{Value}(v) \land \textit{Datatype}(v,t)
\land \textit{IRI}(t)
\\
\forall v \ \textit{Value}(v) & \Rightarrow& v \in\mathbb{S}
\\
\textit{LanguageTag}(v, l) & \equiv& Datatype (v,
\\
&& \texttt{<http://www.w3.org/1999/02/22-rdf-syntax-ns\#langString>} )
\end{eqnarray*}}}

>>>>

The conceptual notion of a \textit{Research Object}
\cite{doi:10.1016/j.future.2011.08.004} is thus realised with the RO-Crate
model and serialised using Linked Data constructs within the RO-Crate
metadata file.

@@

The possibilities of consuming\footnote{Some consideration is needed in processing of RO-Crates as
knowledge graphs, e.g. establishing absolute URIs for files inside a
ZIP archive, detailed in the RO-Crate specification: \url
{https://www.researchobject.org/ro-crate/1.1/appendix/relative-uris.html}.} RO-Crate metadata with such
powerful tools gives another strong reason for using Linked Data as a
foundation. This use of mature Web\footnote{Note that an RO-Crate is not required to be published on the
Web, see Section~\ref{sec:selfdescribed}.} technologies also means its
developers and consumers are not restricted to the Research Object
aspects that have already been specified by the RO-Crate community, but
can extend and integrate RO-Crate in multiple standardised ways.

@@

The RO-Crate specification provides a core set of conventions to
describe research outputs using types and properties applicable across
scientific domains. However we have found that domain-specific use of
RO-Crate will, implicitly or explicitly, form a specialised \textbf{profile}
of RO-Crate; i.e., {a set of conventions, types and properties that are
minimally required and one can expect to be present in that subset of
RO-Crates}. For instance, RO-Crates used for exchange of workflows will
have to contain a data entity of type \texttt{ComputationalWorkflow}, or
cultural heritage records should have a \texttt{contentLocation}.

@@

RO-Crate
mandates\footnote{\url{https://www.researchobject.org/ro-crate/1.1/appendix/jsonld.html}}
the use of flattened, compacted JSON-LD in the RO-Crate Metadata file
\texttt{ro-crate-metadata.json}\footnote{The avid reader may spot that the RO-Crate Metadata file use the
extension \texttt{.json} instead of \texttt{.jsonld}, this is to emphasise the
developer expectations as a JSON format, while the file's JSON-LD
nature is secondary. See \url
{https://github.com/ResearchObject/ro-crate/issues/82}.}
where a single \texttt{@graph} array contains all
the data and contextual entities in a flat list. An example can be seen
in the JSON-LD snippet in Listing \ref{lis1} below, describing a simple RO-Crate
containing data entities described using contextual entities.

@@

In JSON-LD, the \texttt{@context} is a reference to another JSON-LD document
that provides mapping from JSON keys to Linked Data term IRIs, and can
enable various JSON-LD directives to cater for customised JSON
structures for translating to RDF.

@@

The Pacific And Regional Archive for Digital Sources in Endangered
Cultures (PARADISEC\footnote{\url{https://www.paradisec.org.au/}}) \cite{doi:10125/4567}
maintains a repository of more than 500,000 files documenting
endangered languages across more than 16,000 items, collected and
digitised over many years by researchers interviewing and recording
native speakers across the region.

@@

This initially workflow-centric approach was further formalised as the
Wf4Ever Research Object Model \cite{doi:10.1016/j.websem.2015.01.003},
which is a general-purpose research artefact description framework.
This model is based on existing ontologies (FOAF, Dublin Core Terms,
OAI-ORE and AO/OAC precursors to the W3C Web Annotation Model
\cite{Ciccarese:17:WAD}) and adds specializations for workflow models and
executions using W3C PROV-O \cite{PROVO}. The Research Object statements
are saved in a \textit{manifest} (the OAI-ORE \textit{resource map}), with additional
annotation resources containing user-provided details such as title and
description.

We now claim that one barrier for wider adoption of the Wf4Eer Research
Object model for general packaging digital research artefacts was
exactly this re-use of multiple existing vocabularies (FAIR principle
I2: \textit{Metadata use vocabularies that follow FAIR principles}), which in
itself is recognised as a challenge \cite{doi:10.3233/978-1-61499-660-6-9}.
Adapters of the Wf4Ever RO model would have to navigate documentation
of multiple overlapping ontologies, in addition to facing the usual
Semantic Web development choices for RDF serialization formats,
identifier minting and publishing resources on the Web.

@@

\subsection{Example of formalised RO-Crate}

@@

A formalised RO-Crate in $\mathcal{L}_{\mathit{rocrate}}$ can be mapped to
different serializations.
Assume a simplified\footnote{This simplification does not cover the extensive list of literal
datatypes built-in to RDF 1.1, only strings and decimal real numbers.
Likewise, language of literals are not included.} language $\mathcal{L}_{\textit{RDF}}$
based on the RDF abstract syntax \cite{rdfworkinggroup_2014}:
%
{\fontsize{9.8}{11.8}\selectfont{\begin{eqnarray*}
\mathcal{L}_{\textit{RDF}} & \equiv& \bigl\{ \textit{Triple}(s,p,o),
\textit{IRI}(i), \textit{BlankNode}(b), \textit{Literal}(s), \mathbb{IRI},
\mathbb{S}, \mathbb{R} \bigr\}
\\
\mathbb{D}_{\textit{RDF}} & \equiv& \mathbb{S}
\\
\forall i \ \textit{IRI}(i) & \Rightarrow& i \in\mathbb{IRI}
\\
\forall s \forall p \forall o \ \textit{Triple}(s,p,o) & \Rightarrow& \bigl(
\textit{IRI}(s) \lor \textit{BlankNode}(s) \bigr) \land
\\
& & \textit{IRI}(p) \land
\\
& & \bigl(\textit{IRI}(o) \lor \textit{BlankNode}(o) \lor \textit{Literal}(o)
\bigr)
\\
\textit{Literal}(v) & \models& \textit{Value}(v) \land \textit{Datatype}(v,t)
\land \textit{IRI}(t)
\\
\forall v \ \textit{Value}(v) & \Rightarrow& v \in\mathbb{S}
\\
\textit{LanguageTag}(v, l) & \equiv& Datatype (v,
\\
&& \texttt{<http://www.w3.org/1999/02/22-rdf-syntax-ns\#langString>} )
\end{eqnarray*}}}
